\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-10-23 paper.tex
Update

Any eARG can be converted to a sample-resolved gARG using the same machinery,
and the two-step process is illustrated in Fig.~\ref{fig-ancestry-resolution}.
The first step is to take the input eARG (Fig.~\ref{fig-ancestry-resolution}A)
and convert all event nodes to genome nodes and add inheritance annotations to
the gARG's edges (Fig.~\ref{fig-ancestry-resolution}B).
If the node is a common ancestor node, we annotate the single
outbound edge with the interval $[0,L)$, for a genome of length $L$. If the
node is a recombination event with a breakpoint $x$, we annotate the two
outbound edges respectively with the intervals $[0, x)$ and $[x, L)$. These
inheritance interval annotations are clearly in one-to-one correspondence with
the information in the input eARG. They are also analogous to the
inheritance intervals we get on the edges in a prospective ARG
produced by a forwards-time simulation, which are concerned with recording
the local genetic relationship between a parent and child genome and not
necessarily minimal in terms of the resolved ancestral material of a sample.
Thus, the final step is to use the ``simplify'' algorithm to perform this
resolution~(Fig.\ref{fig-ancestry-resolution}C).

The sample-resolved gARG of Fig.~\ref{fig-ancestry-resolution}C
differs in some important ways to the
original eARG (Fig.~\ref{fig-ancestry-resolution}A).
Firstly, we can see that some nodes and edges have been removed entirely
from the graph.
The ``grand MRCA'' \noderef{q} is omitted from the
sample-resolved gARG because all segments of the genome have
fully coalesced before it is reached. Likewise, the edge
between \noderef{g} and \noderef{j} is omitted because the recombination
event at position $5$ (represented by node \noderef{g})
fell in non-ancestral material.
More generally, we can see that the sample resolved
gARG of Fig.~\ref{fig-ancestry-resolution}C
allows for ``local'' inspection
of an ARG in a way that is not possible by storing
common ancestor and recombination events in an eARG. Because
the ancestral material is stored with each edge, the
cumulative effects of events over time can be reasoned
about, without first ``replaying'' those events. Many computations
that we wish to perform on an ARG will require resolving
the ancestral material with respect to a sample.
% this is weak, but would be good to emphasise this point.
% you just end up running simplify lots of times, in your
% downstream algs
The gARG encoding
allows us to perform this resolution of ancestral material once
and to store the result,
whereas the eARG encoding requires us to repeat the process
each time.

@@

\section{ARGs and local trees}
\label{sec-ARG-and-local-trees}
Local trees, describing the genetic genealogy of a sample
at a particular location along the genome, are
embedded within an ARG, and the relationship
between the two is subtle and important.
Note that many authors use the term ``marginal''
% could list loads here, but not much point.
trees~\citep[e.g.][]{griffiths1996ancestral,minichiello2006mapping,kelleher2016efficient}
which we avoid due to the potential confusion with the
statistical concept of marginal distributions
(potentially leading to interpretation as an integrated tree over many positions).

% TODO Need to make clear here what came from where. Basic combinatorial object
% cite Knuth. Application to genetic ancestry, sim papers. Need to look
% up what's said in favour of this integer based approach in which paper
% and refer to tat.
To make the following discussions precise, we provide
a concrete definition and explicit process for local tree extraction.
We can describe a local tree as an
oriented tree~\cite[p.\ 461]{knuth11combinatorial},
a sequence of integers $\pi_1, \pi_2\dots$ where $\pi_u$ denotes the parent of
node $u$~\citep{kelleher2013coalescent,kelleher2014coalescent,kelleher2016efficient}.
If $u$ is a root
% I'm not sure if this reachability stuff is a helpful clarification, and whether it'll just
% confuse people. It depends on some tricky details.
(or not reachable from the samples)
in the local tree at position $x$, then we adopt the convention that $\pi_u = -1$.
% TODO Give example of an oriented tree from one of the figs?
% TODO
% A node in an oriented tree can have
% zero, one,
% or many children.

To recover the local tree $\pi$ at position $x$, we begin by
setting $\pi_u = -1$ for each $u\in N$. Then, for each sample
node in $S$ we trace its path rootwards through the
ARG for position $x$, and record this path in $\pi$.
Specifically, at a given node $u$,
we find an edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$, and set
$\pi_c \leftarrow p$. We then set $u \leftarrow p$, and repeat
until either $\pi_u \neq -1$ (indicating we have traversed this section
of the ARG already on the path from another sample) or there
is no matching outbound edge (indicating we are at a root).

The nodes in a local tree encoded
in this way are \emph{labelled} (i.e.\ carry the identity
of the corresponding ARG node), and this is fundamentally
important to how a local tree relates to its ARG.
If we are given the sequence of local trees for a gARG
encoded as an oriented tree (or in any other form in which
the tree nodes are explicitly labelled with the identity of the
corresponding ARG node),
along with the genome interval covered by each tree,
then we can recover precisely the same ARG. Thus, under this
definition, there
is a one-to-one correspondence between an ARG and
the sequence of local trees that it encodes.

This is not the prevailing view, however.
\cite{kuhner2017consensus} argue that the
``interval-tree'' representation of
an ARG (the local trees and the genome intervals they cover)
``does not contain all of the information in the underlying ARG: it lacks the
number of recombinations occuring at each site, the times at which
recombinations occurred, and the specific sequences involved as recombination
partners.''
\cite{shipilina2023origin} discuss the same ideas,
and note that the
``full ARG...~contains more information than the series of tree
sequences along the genome''.
These statements that an ARG
contains more information than its local trees is only true
if we use a representation of the local trees that discards
information about those local trees that can be derived from the ARG.
The internal nodes of evolutionary trees are usually
considered to be \emph{unlabelled}
because they most often correspond
to anonymous hypothetical ancestors.
Internal ARG nodes may likewise correspond to anonymous hypothetical
ancestors, but if we do not label the tree nodes we lose the fact that
the \emph{same} ancestors persist across multiple trees.
Besides having labelled nodes, the local trees must contain
unary nodes marking recombinations
and other processes that lead to lineages passing through a given
ARG node without coalescence in the local trees
(see Section~\ref{sec-locally-unary-edges}).
If the local trees contain labelled internal nodes, they precisely
describe the corresponding path through the graph for each
genome interval and there is exactly as
much information in them as in an ARG.

>>>>

Any eARG can be converted to a sample-resolved gARG using the same machinery,
and the two-step process is illustrated in Fig.~\ref{fig-ancestry-resolution}.
The first step is to take the input eARG (Fig.~\ref{fig-ancestry-resolution}A),
duplicate its graph topology, and then add inheritance annotations
to each of the gARG's edges (Fig.~\ref{fig-ancestry-resolution}B).
If a given node is a common ancestor event, we annotate the single
outbound edge with the interval $[0,L)$, for a genome of length $L$. If the
node is a recombination event with a breakpoint $x$, we annotate the two
outbound edges respectively with the intervals $[0, x)$ and $[x, L)$. These
inheritance interval annotations are clearly in one-to-one correspondence with
the information in the input eARG. They are also analogous to the
inheritance intervals we get on the edges in a prospective gARG
produced by a forwards-time simulation, which are concerned with recording
the local genetic relationship between a parent and child genome and not
necessarily minimal in terms of the resolved ancestral material of a sample.
Thus, the final step is to use the ``simplify'' algorithm to perform this
resolution (Fig.~\ref{fig-ancestry-resolution}C).

The sample-resolved gARG of Fig.~\ref{fig-ancestry-resolution}C
differs in some important ways to the
original eARG (Fig.~\ref{fig-ancestry-resolution}A).
Firstly, we can see that some nodes and edges have been removed entirely
from the graph.
The ``grand MRCA'' \noderef{q} is omitted from the
sample-resolved gARG because all segments of the genome have
fully coalesced before it is reached. Likewise, the edge
between \noderef{g} and \noderef{j} is omitted because the recombination
event at position $5$ (represented by node \noderef{g})
fell in non-ancestral material.
More generally, we can see that the sample resolved
gARG of Fig.~\ref{fig-ancestry-resolution}C
allows for ``local'' inspection
of an ARG in a way that is not possible by storing
common ancestor and recombination events in an eARG. Because
the ancestral material is stored with each edge, the
cumulative effects of events over time can be reasoned
about, without first ``replaying'' those events.
Many computations
that we wish to perform on an ARG will require resolving
the ancestral material with respect to a sample.
% this is weak, but would be good to emphasise this point.
% you just end up running simplify lots of times, in your
% downstream algs
The gARG encoding
allows us to perform this once
and to store the result,
whereas the eARG encoding requires us to repeat the process
each time.

@@

\section{ARGs and local trees}
\label{sec-ARG-and-local-trees}
% (We use ``local'' rather than
% ``marginal''
% % could list loads here, but not much point.
% trees~\citep[e.g.][]{griffiths1996ancestral,minichiello2006mapping,kelleher2016efficient}
% to avoid the potential confusion with the
% statistical concept of marginal distributions.)
A fundamental property of genetics is that a
% TODO is this (genome) clarification helpful or confusing?
given DNA nucleotide is inherited from exactly one parent (genome),
both at an organismal and cell-by-cell level
(Appendix~\ref{sec-cell-lineages-and-args}).
These chains of single-parent inheritance over time give rise,
by definition, to a tree structure.
As a result of recombination, adjacent nucleotides can have
different chains of inheritance, and
an ARG encodes the entire ensemble of local trees along the
genome for a given set of sample genomes.
The relationship between an ARG and its corresponding
local trees is subtle and important. To make the following
discussions precise, we will define a concrete combinatorial encoding
for local trees and an explicit procedure for constructing them.

Oriented trees provide a
convenient formalism to capture these relationships in a well-defined
combinatorial object.
% There's no need for a comma
Let $\pi_1\dots\pi_n$ be a sequence of integers, such that $\pi_u$
denotes the parent of node $u$, and $\pi_u = 0$ if $u$ is a
root~\cite[p.\ 461]{knuth11combinatorial}.
This encoding is particularly useful to describe
evolutionary trees because parent-child relationships are
important but the ordering of children at a node is
not~\citep{kelleher2013coalescent,kelleher2014coalescent,
kelleher2016efficient}.
Thus, for a given gARG with nodes $\{1, \dots, n\}$ and
edges $E$, we recover the local tree as follows.
We begin by
setting $\pi_u = 0$ for each $1 \leq u \leq n$. Then, for each sample
node in $S$ we trace its path rootwards through the
ARG for position $x$, and record this path in $\pi$.
Specifically, at a given node $u$,
we find an edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$, and set
$\pi_c \leftarrow p$. We then set $u \leftarrow p$, and repeat
until either $\pi_u \neq 0$ (indicating we have traversed this section
of the ARG already on the path from another sample) or there
is no matching outbound edge (indicating we are at a root).
Note that the trees for an ARG are ``sparse''~\citep{kelleher2016efficient},
because many ancestral nodes will not be reachable from the
samples at a given position.

This combinatorial approach provides at least
one novel insight into the relationship between ARGs and local trees.
If we are given the sequence of local trees for a gARG
encoded as an oriented tree along with the genome interval
covered by each tree, then we can recover precisely the same ARG.
% I guess we should be a bit more formal here?
Thus, under this definition, there
is a one-to-one correspondence between an ARG and
the sequence of local trees that it encodes.

This is not the prevailing view, however.
\cite{kuhner2017consensus} argue that the
``interval-tree'' representation of
an ARG (the local trees and the genome intervals they cover)
``does not contain all of the information in the underlying ARG: it lacks the
number of recombinations occuring at each site, the times at which
recombinations occurred, and the specific sequences involved as recombination
partners.''
\cite{shipilina2023origin} discuss the same ideas,
and note that the
``full ARG...~contains more information than the series of tree
sequences along the genome''.
% Could cite some more? E.g. KwARG paper says
% "We note that ARGs contain more information than local trees"
% If even Ana and Jotun are wrong about this it then it must be
% significant! ;)
These statements that an ARG
contains more information than its local trees is only true
if we use a representation of the local trees that discards
information about those local trees that can be derived from the ARG.

There are two properties of how evolutionary trees are conventionally
represented that lead to this
% Yes, OK, we can rephrase. It is a misunderstanding though
misunderstanding of the relationship between local trees and an ARG.
Firstly, the internal nodes of evolutionary trees are usually
considered to be \emph{unlabelled}.
Defining ARG nodes as integers and using the oriented tree encoding
explicitly labels internal nodes, and makes the relationship between
tree and ARG nodes explicit.

[WIP]

because they most often correspond
to anonymous hypothetical ancestors.
Internal ARG nodes may likewise correspond to anonymous hypothetical
ancestors, but if we do not label the tree nodes we lose the fact that
the \emph{same} ancestors persist across multiple trees.
Besides having labelled nodes, the local trees must contain
unary nodes marking recombinations
and other processes that lead to lineages passing through a given
ARG node without coalescence in the local trees
(see Section~\ref{sec-locally-unary-edges}).
If the local trees contain labelled internal nodes, they precisely
describe the corresponding path through the graph for each
genome interval and there is exactly as
much information in them as in an ARG.
